---
title: "General Society Survey Data Analysis"
author: "Arindam Samanta"
date: January 19 2020
output:
  
  word_document: default
---

```{r setup,echo=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```
## Part 1&2: Is there a significant relationship between the number of siblings a survey respondent has and number of his or her children?”

### General Society Survey data 

Data for this assignment originated from the General Society Survey (GSS). The GSS gathers data on contemporary American society in order to monitor and explain trends and constants in attitudes, behaviors, and attributes. Hundreds of trends have been tracked since 1972. In addition, since the GSS adopted questions from earlier surveys, trends can be followed for up to 70 years. Only data from the 2016 GSS survey is included in this dataset [gss2016](http://content.bellevue.edu/cst/dsc/520/id/resources/gss-2016.csv).

## Preparations 
We have used the following libraries for our regression analysis. Regression analysis is used in stats to find trends in data. For example, we are guessing  that there is a connection between how many child a respondant has and how many siblings he or she has; regression analysis can help us quantify that.
```{r loading_packages, echo = TRUE, results='hide', warning=FALSE, message=FALSE}
library(dplyr)
library(ggplot2)
library(readr)
library(corrplot)
```
## Loading data to include the SIBS,CHILDS and SEX variables from the actual data set
As we are interested here only in the above features in our data set.
```{r , echo=TRUE, results='hide'}
#loading data into dataframe
mycols <- cols_only(
                  CHILDS = col_double(),
                  SIBS = col_double(),
                  SEX = col_factor()
                  )
wd <- getwd()
fname <- "gss-2016.csv"
path_to_file <- paste(wd,'/dataset/',fname, sep = "")
path_to_file

survey_data <- read_csv(path_to_file)

gss2016.sub <- read_csv(path_to_file, 
                        col_types = mycols)
```
For this analysis we are interested in looking at the GSS 2016 survey data, specifically the Siblings and Childs variables. 

```{r }
# looking at the first few rows
head(gss2016.sub)
```

Summaize the data based on CHILDS to see how the distribution looks

```{r}
## Summarizing the data set by # CHILDS
gss2016.sub %>%
  group_by(CHILDS) %>%
  summarize(n())

```
Summaize the data based on SIBS to see how the distribution looks

```{r}
## Summarizing the data set by # SIBLINGS
gss2016.sub %>%
  group_by(SIBS) %>%
  summarize(n())

```
The summarized data shows that there are 8 rows that has Child as NA and 5 rows that has SIBS as NA.

### Correlation Coefficient (Pearson Product moment Correlation)
The covarience between the two variables SIBS and CHILD is `r gss2016.sub %>% summarize(cor(gss2016.sub$CHILDS,gss2016.sub$SIBS, use = "pairwise.complete.obs")) `

The relationship shows a weak positive trend,irrespective on how we measure the variables which means that No. of siblings do not have any relationship to the number of child a respondant has. This could also be seen in the plot below and fitting a line.
## Including Plots
Now Construct a scatterplot of these two variables in R studio and place the best-fit linear regression line on the scatterplot. Describe the relationship between the number of siblings a respondent has (SIBS) and the number of his or her children (CHILDS).

```{r , echo=TRUE, warning=FALSE, message=FALSE}
# Creating a scatter plot with the survey data
ggplot(data = gss2016.sub, aes(x = SIBS, y = CHILDS)) +
  
  geom_point(position = "jitter") +
  geom_smooth(method = "lm", se = FALSE)
```

## Expanding the analysis to bring in the third variable

Generating a scatter plot with the third variable SEX of the respondant.

```{r , echo=TRUE, warning=FALSE, message=FALSE}
# Creating a scatter plot with the third variable SEX survey data
ggplot(data = gss2016.sub, aes(x = SIBS, y = CHILDS, color = SEX)) +
  
  geom_point(position = "jitter") +
  geom_smooth(method = "lm", se = FALSE)
```

## Trying to fit a linear model
In this model we are trying to see if the # of SIBS could predict the # of CHILD.The intercept and the slope of the model is as below. Given below is the coefficient values and summary of the model.

```{r}
# The formula for the linear model
fmla <- CHILDS ~ SIBS
# Fitting the formula to a model
mod <- lm(fmla, data = gss2016.sub)
# Coeeficient of the model
 coef(mod)
 summary(mod)
```

Trying to predict using the model. 

```{r, echo=TRUE, results='hide'}
# Created a new dataframe with 3 SIBS
new_df3 <- data.frame("SIBS" = 3 )
# Predicted no. of CHILDS
pred3 <- predict(mod, newdata = new_df3)
pred3

# Created a new dataframe with 0 SIBS
new_df0 <- data.frame("SIBS" = 0 )
# Predicted no. of CHILDS
pred0 <- predict(mod, newdata = new_df0)
pred0

```
Predicted number of children for someone with three siblings :`r pred3`

Predicted number of children for someone with no siblings :`r pred0`